Ethnicity sensitive author disambiguation using semi-supervised learning

### 2016-04-24 paper/author_disambiguation.tex
kesw: fits in 15 pages.
* Removed table with maximum blocking recall
* Removed few citations to unsupervised clustering
* Shrunk the rfe figure

Signed-off-by: Mateusz Susik <mateuszsusik@gmail.com>

\usepackage{xspace}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage[hidelinks]{hyperref}
\newcommand*{\eg}{e.g.\@\xspace}
\newcommand*{\ie}{i.e.\@\xspace}
\usepackage{color}
\newcommand{\msnote}[1]{\textcolor{blue}{[MS: #1]}}

@@

\authorrunning{Gilles Louppe et al.}
\author{Gilles Louppe\inst{1}\and Hussein T. Al-Natsheh\inst{2}\and Mateusz Susik\inst{3}\and\\ 
Eamonn James Maguire\inst{1}}

@@

Efforts and solutions to author disambiguation have been proposed from various
communities \cite{liu2014author}. On the one hand, libraries have maintained
authorship control through manual curation, either in a centralized way by
hiring professional collaborators or through developing services that invite authors to register
their publications themselves (\eg, Google Scholar or Inspire-HEP).
Recent efforts to create persistent digital identifiers assigned to researchers (\eg, ORCID or ResearcherID),
with the objective to embed these identifiers in the submission workflow of publishers
or repositories (\eg, Elsevier, arXiv or Inspire-HEP), would univocally solve any disambiguation issue.
As the centralized manual authorship control is expensive and the success of persistent digital
identifiers requires large and ubiquitous adoption by both researchers and publishers, fully
automated machine learning-based methods have been proposed during the past decade to
provide immediate, less costly, and satisfactory solutions to author
disambiguation. In this work, our goal is to explore and demonstrate how both
approaches can coexist and benefit from each other.
In particular, we study how labeled data obtained through manual curation (either centralized or
crowdsourced) can be exploited (i) to learn an accurate classifier for
identifying coreferring authors, and (ii) to guide the clustering of scientific
publications by distinct authors in a semi-supervised way.
Our analysis of parameters and features of this large dataset reveal that the general pipeline
commonly used in existing solutions is an effective approach for author disambiguation.
Additionally, we propose alternative strategies for blocking based on the
phonetization of author names to increase recall and ethnicity-sensitive features for learning a
linkage function which tailor our author disambiguation to non-Western author names whenever
necessary.

@@

Training data is usually not easily available, therefore unsupervised approaches propose
the use of domain-specific, manually designed, linkage functions tailored towards author
disambiguation \cite{malin2005unsupervised,mcrae2006also,song2007efficient,
soler2007separating, kang2009co,fan2011graph,schulz2014exploiting}.
These approaches have the advantage of not requiring hand-labeled data, but generally do
not perform as well as supervised approaches.
To reconcile both worlds, semi-supervised methods make use of small, manually verified, clusters of
publications and/or high-precision domain-specific rules to build a training
set of pairs of publications, from which a linkage function is then built using supervised learning
\cite{ferreira2010effective,torvik2009author,levin2012citation}.

@@

In this
framework, author disambiguation can be stated as the problem of
finding a partition ${\cal C} = \{ c_0, ..., c_{M-1} \}$ of ${\cal S}$ such
that ${\cal S} = \cup_{i=0}^{M-1} c_i$, $c_i \cap c_j = \phi$ for all $i \neq
j$, and where subsets $c_i$, or clusters, each corresponds to the set of all
signatures belonging to the same individual $a_i$. Alternatively, the set
${\cal A}$ may remain (possibly partially) unknown, such that author
disambiguation boils down to finding a partition ${\cal C}$ where
subsets $c_i$  each correspond to the set of all signatures from the same
individual (without knowing who). Finally, in the case of partially annotated databases as studied in
this work, the set extends with the partial knowledge ${\cal C}^\prime = \{ c_0^\prime, ..., c_{M-1}^\prime \}$ of ${\cal C}$,
such that $c_i^\prime \subseteq c_i$, where $c_i^\prime$ may be empty.
Or put otherwise, the set extends with the assumption that all signatures
$s \in c_i^\prime$ belong to the same author.

Our algorithm is composed of three parts, as sketched in Figure~\ref{fig:workflow}: (i) a blocking
scheme whose goal is to roughly pre-cluster signatures ${\cal S}$ into smaller groups in order to
reduce computational complexity; (ii) the construction of a linkage function
$d$ between signatures using supervised learning; and (iii) the
semi-supervised clustering of all signatures within the same block, using $d$ as a pseudo distance metric.

@@

As in previous works, the first part of our algorithm consists of dividing
signatures ${\cal S}$ into disjoint subsets ${\cal S}_{b_0}, ..., {\cal
S}_{b_{K-1}}$, or \textit{blocks} \cite{fellegi69}, followed by carrying out
author disambiguation on each one of these blocks independently.
By doing so, the computational complexity of clustering (see Section~\ref{methods:clustering})
typically reduces from $O(|{\cal S}|^2)$ to $O(\sum_b |{\cal S}_b|^2)$, which is much more
tractable as the number of signatures increases.
Since disambiguation is performed independently per block, a good blocking strategy should be
designed such that signatures from the same author are all mapped to the same
block, otherwise their correct clustering would not be possible in later stages of the workflow.
As a result, blocking should be a balance between reduced complexity and maximum recall.

The simplest and most common strategy for blocking, referred to hereon in as \textit{Surname and First Initial (SFI)},
groups signatures together if they share the same surname(s) and the same first
given name initial (\eg, \emph{SFI}("Doe, John") $==$ "Doe, J").
Despite satisfactory performance, there are several cases where this simple strategy fails to cluster related pairs of signatures together, including:

@@

\textit{Supervised classification.} The second part of the algorithm is the
automatic construction of a pair-wise linkage function between signatures for use
during the clustering step which groups all signatures from the same author.

Formally, the goal is to build a function $d: {\cal S} \times {\cal S} \mapsto
[0, 1]$, such that $d(s_1, s_2)$ approaches $0$ if both signatures $s_1$ and
$s_2$ belong to the same author, and $1$ otherwise.
This problem can be cast as a standard supervised classification task, where inputs
are pairs of signatures and outputs are classes $0$ (same authors), and $1$
(distinct authors). In this work, we evaluate Random Forests (RF, \cite{breiman2001random}),
Gradient Boosted Regression Trees (GBRT, \cite{friedman2001greedy}),
and Logistic Regression \cite{fan2008liblinear} as classifiers.

\textit{Input features.} In most cases, supervised learning algorithms assume
the input space ${\cal X}$ to be numeric (\eg, $\mathbb{R}^p$), making them
not directly applicable to structured input spaces such as ${\cal S} \times
{\cal S}$.
Following previous works, pairs of signatures $(s_1, s_2)$ are first transformed to vectors $v \in \mathbb{R}^p$
by building so-called similarity profiles \cite{treeratpituk2009disambiguating} on which supervised learning is carried out.
In this work, we design and evaluate fifteen standard input
features based on the comparison of signature fields, as reported in the first
half of Table~\ref{table:features}.
As an illustrative example, the \textit{Full name} feature corresponds to the similarity between the (full)
author name fields of the two signatures, as measured using as combination
operator the cosine similarity between their respective $(n,m)$-\emph{TF-IDF} vector
representations\footnote{$(n,m)$ denotes that the \emph{TF-IDF} vectors are computed
from character $n$, $n+1$, ..., $m$-grams.
When not specified, \emph{TF-IDF} vectors are otherwise computed from words.}.

@@

\subsection{Data}

The author disambiguation solution proposed in this work, along with its
enhancements, are evaluated on data extracted from the \emph{INSPIRE} portal
\cite{gentil2009information}, a digital library for scientific literature in
high-energy physics.
Overall, the portal holds more than 1 million publications ${\cal P}$,
forming in total a set ${\cal S}$ of more than 10 million signatures.
Out of these, around 13\% have been \textit{claimed} by their
original authors, marked as such by professional curators or automatically assigned to their true authors thanks
to persistent identifiers provided by publishers or other sources.
Together, they constitute a trusted set $({\cal S}^\prime, {\cal C}^\prime)$ of 15388 distinct individuals sharing
36340 unique author names spread within 1201763 signatures on 360066
publications. This data covers several decades in time and dozens of author
nationalities worldwide.

@@

As commonly performed in author disambiguation research,
we evaluate the predicted clusters over testing data  ${\cal C}_\text{test}^\prime$,
using both B3 and pairwise precision, recall
and F-measure, as defined below:
\begin{small}
\begin{align}
P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{1}{|{\cal S}|} \sum_{s \in {\cal S}} \frac{|c(s) \cap \widehat{c}(s)|}{|\widehat{c}(s)|} \\
R_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{1}{|{\cal S}|} \sum_{s \in {\cal S}} \frac{|c(s) \cap \widehat{c}(s)|}{|c(s)|}\\
F_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{2 P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) R_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S})}{P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) + P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S})} \\
P_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{|p({\cal C}) \cap p(\widehat{\cal C})|}{|p(\widehat{\cal C})|}\\
R_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{|p({\cal C}) \cap p(\widehat{\cal C})|}{|p({\cal C})|}\\
F_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{2 P_\text{pairwise}({\cal C}, \widehat{\cal C}) R_\text{pairwise}({\cal C}, \widehat{\cal C})}{P_\text{pairwise}({\cal C}, \widehat{\cal C}) + R_\text{pairwise}({\cal C}, \widehat{\cal C})}
\end{align}
\end{small}
and where $c(s)$ (resp. $\widehat{c}(s)$) is the cluster $c \in {\cal C}$ such that
$s \in c$ (resp. the cluster $\widehat{c} \in \widehat{\cal C}$ such that $s
\in \widehat{c}$), and where $p({\cal C}) = \cup_{c \in {\cal C}} \{ (s_1, s_2)
| s_1, s_2 \in c, s_1 \neq s_2 \}$ is the set of all pairs of signatures from
the same clusters in ${\cal C}$.
The F-measure is the harmonic mean between these two quantities.
In the analysis below, we rely primarily on the B3 F-measure for discussing results, as the pairwise variant
tends to favor large clusters (because the number of pairs is quadratic with the cluster size),
hence unfairly giving preference to authors with many publications.
By contrast, the B3 F-measure weights clusters linearly with respect to their size.
General conclusions drawn below remain however consistent for pairwise F.

@@

\textit{Baseline.} The simplest baseline against which we compare our results
consists in grouping all signatures sharing the same (normalized) surname(s) and
the same (normalized) first given name initial. As it does not require the usage
of any machine learning component, this baseline provides a simple and fast
solution yielding decent results, as reported at the top of
Table~\ref{table:results}

@@

\begin{itemize}
  \item Blocking: same surname and the same first given name initial strategy (SFI) and with name normalization;
  \item Linkage function: all 22 features defined in Table~\ref{table:features},
  gradient boosted regression trees as supervised learning algorithm
  and a training set of pairs built from $({\cal S}_\text{train}^\prime, {\cal C}_\text{train}^\prime)$, by balancing easy and difficult cases.
  \item Clustering: agglomerative clustering using average linkage and
  block cuts found to maximize \\ $F_\text{B3}({\cal C}_\text{train}^\prime, \widehat{\cal C}_\text{train}^\prime, {\cal S}_\text{train}^\prime)$.
\end{itemize}

@@

Finally, as discussed previously, the seemingly insignificant step of
normalizing author names (stripping accents, removing affixes), as performed in the
state-of-the-art, is shown to be important. Results from Table~\ref{table:results} clearly suggest that not
normalizing significantly reduces performance (yielding an F-measure of $0.9830$ when normalizing,
but decreasing to $0.9791$ when raw author name strings are used instead).

>>>>

\usepackage{xspace}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{cite}
\usepackage[hidelinks]{hyperref}
\newcommand*{\eg}{e.g.\@\xspace}
\newcommand*{\ie}{i.e.\@\xspace}
\usepackage{color}
\setlength{\intextsep}{10pt plus 2pt minus 2pt}
\newcommand{\msnote}[1]{\textcolor{blue}{[MS: #1]}}

@@

\authorrunning{G. Louppe, H. Al-Natsheh, M. Susik, E. Maguire}
\author{Gilles Louppe\inst{1}\and Hussein T. Al-Natsheh\inst{2}\and Mateusz Susik\inst{3}\and\\ 
Eamonn James Maguire\inst{1}}

@@

Solutions to author disambiguation have been proposed from various
communities \cite{liu2014author}. On the one hand, libraries have maintained
authorship control through manual curation, either in a centralized way by
hiring professional collaborators or through developing services that invite authors to register
their publications themselves (\eg, Google Scholar or Inspire-HEP).
Recent efforts to create persistent digital identifiers assigned to researchers (\eg, ORCID or ResearcherID),
with the objective to embed these identifiers in the submission workflow of publishers
or repositories (\eg, Elsevier, arXiv or Inspire-HEP), would univocally solve any disambiguation issue.
As the centralized manual authorship control is expensive and the success of persistent digital
identifiers requires large and ubiquitous adoption by both researchers and publishers, fully
automated machine learning-based methods have been proposed to
provide immediate, less costly, and satisfactory solutions to author
disambiguation. In this work, we study how labeled data obtained through manual curation (either centralized or
crowdsourced) can be exploited (i) to learn an accurate classifier for
identifying coreferring authors, and (ii) to guide the clustering of scientific
publications by distinct authors in a semi-supervised way.
Our analysis of parameters and features of this large dataset reveal that the general pipeline
commonly used in existing solutions is an effective approach for author disambiguation.
Moreover, we propose better strategies for blocking based on the
phonetization of author names to increase recall and ethnicity-sensitive features for learning a
linkage function which tailor our author disambiguation to non-Western author names.

@@

Training data is usually not easily available, therefore unsupervised approaches propose
the use of domain-specific, manually designed, linkage functions tailored towards author
disambiguation \cite{malin2005unsupervised,song2007efficient, kang2009co,schulz2014exploiting}.
These approaches have the advantage of not requiring hand-labeled data, but generally do
not perform as well as supervised approaches.
To reconcile both worlds, semi-supervised methods make use of small, manually verified clusters of
publications and/or high-precision domain-specific rules to build a training
set of pairs of publications, from which a linkage function is then built using supervised learning
\cite{ferreira2010effective,torvik2009author,levin2012citation}.

@@

Author disambiguation can be stated as the problem of
finding a partition ${\cal C} = \{ c_0, ..., c_{M-1} \}$ of ${\cal S}$ such
that ${\cal S} = \cup_{i=0}^{M-1} c_i$, $c_i \cap c_j = \phi$ for all $i \neq
j$, and where subsets $c_i$, or clusters, each corresponds to the set of all
signatures belonging to the same individual $a_i$. Alternatively, the set
${\cal A}$ may remain (possibly partially) unknown, such that author
disambiguation boils down to finding a partition ${\cal C}$ where
subsets $c_i$  each correspond to the set of all signatures from the same
individual (without knowing who). Finally, in the case of partially annotated databases as studied in
this work, the set extends with the partial knowledge ${\cal C}^\prime = \{ c_0^\prime, ..., c_{M-1}^\prime \}$ of ${\cal C}$,
such that $c_i^\prime \subseteq c_i$, where $c_i^\prime$ may be empty.

Our algorithm is composed of three parts (Figure~\ref{fig:workflow}): (i) a blocking
scheme whose goal is to pre-cluster signatures ${\cal S}$ into smaller groups; (ii) the construction of a linkage function
$d$ between signatures using supervised learning; and (iii) the
semi-supervised clustering of all signatures within the same block, using $d$ as a pseudo distance metric.

@@

As in previous works, the first part of our algorithm consists of dividing
signatures ${\cal S}$ into disjoint subsets ${\cal S}_{b_0}, ..., {\cal
S}_{b_{K-1}}$, or \textit{blocks} \cite{fellegi69}, followed by carrying out
author disambiguation on each one of these blocks independently.
By doing so, the computational complexity of clustering (see Section~\ref{methods:clustering})
typically reduces from $O(|{\cal S}|^2)$ to $O(\sum_b |{\cal S}_b|^2)$.
Since disambiguation is performed independently per block, a good blocking strategy should be
designed such that signatures from the same author are all mapped to the same
block, otherwise their correct clustering would not be possible in later stages of the workflow.
As a result, blocking should be a balance between reduced complexity and maximum recall.

The simplest and most common strategy for blocking, referred to hereon in as \textit{Surname and First Initial (SFI)},
groups signatures together if they share the same surname(s) and the same first
given name initial.
Despite satisfactory performance, there are several cases where this simple strategy fails to cluster related pairs of signatures together, including:

@@

\subsubsection{Supervised classification.} The second part of the algorithm is the
automatic construction of a pair-wise linkage function between signatures for use
during the clustering step which groups all signatures from the same author.

Formally, the goal is to build a function $d: {\cal S} \times {\cal S} \mapsto
[0, 1]$, such that $d(s_1, s_2)$ approaches $0$ if both signatures $s_1$ and
$s_2$ belong to the same author, and $1$ otherwise.
This problem can be cast as a supervised classification task, where inputs
are pairs of signatures and outputs are classes $0$ (same authors), and $1$
(distinct authors). In this work, we evaluate Random Forests (RF, \cite{breiman2001random}),
Gradient Boosted Regression Trees (GBRT, \cite{friedman2001greedy}),
and Logistic Regression \cite{fan2008liblinear} as classifiers.

%In most cases, supervised learning algorithms assume
%the input space ${\cal X}$ to be numeric (\eg, $\mathbb{R}^p$), making them
%not directly applicable to structured input spaces such as ${\cal S} \times
%{\cal S}$.

\subsubsection{Input features.} Following previous works, pairs of signatures $(s_1, s_2)$ are first transformed to vectors $v \in \mathbb{R}^p$
by building so-called similarity profiles \cite{treeratpituk2009disambiguating} on which supervised learning is carried out.
In this work, we design and evaluate fifteen standard input
features based on the comparison of signature fields, as reported in the first
half of Table~\ref{table:features}.
As an illustrative example, the \textit{Full name} feature corresponds to the similarity between the (full)
author name fields of the two signatures, as measured using as combination
operator the cosine similarity between their respective $(n,m)$-\emph{TF-IDF} vector
representations\footnote{$(n,m)$ denotes that the \emph{TF-IDF} vectors are computed
from character $n$, $n+1$, ..., $m$-grams.
When not specified, \emph{TF-IDF} vectors are otherwise computed from words.}.

@@

All the solutions proposed in this work sare evaluated on data extracted from
the \emph{INSPIRE} portal \cite{gentil2009information}, a digital library for scientific literature in
high-energy physics.
Overall, the portal holds more than 1 million publications ${\cal P}$,
forming in total a set ${\cal S}$ of more than 10 million signatures.
Out of these, around 13\% have been \textit{claimed} by their
original authors, marked as such by professional curators or automatically assigned to their true authors thanks
to persistent identifiers provided by publishers or other sources.
Together, they constitute a trusted set $({\cal S}^\prime, {\cal C}^\prime)$ of 15388 distinct individuals sharing
36340 unique author names spread within 1201763 signatures on 360066
publications. This data covers several decades in time and dozens of author
nationalities worldwide.

@@

As commonly performed in author disambiguation research,
we evaluate the predicted clusters over testing data  ${\cal C}_\text{test}^\prime$,
using both B3 and pairwise precision, recall
and F-measure, as defined below:
\begin{small}
\begin{align}
P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{1}{|{\cal S}|} \sum_{s \in {\cal S}} \frac{|c(s) \cap \widehat{c}(s)|}{|\widehat{c}(s)|} \qquad
R_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) = \frac{1}{|{\cal S}|} \sum_{s \in {\cal S}} \frac{|c(s) \cap \widehat{c}(s)|}{|c(s)|}\\
F_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) &= \frac{2 P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) R_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S})}{P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S}) + P_\text{B3}({\cal C}, \widehat{\cal C}, {\cal S})} \\
P_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{|p({\cal C}) \cap p(\widehat{\cal C})|}{|p(\widehat{\cal C})|} \qquad \qquad
R_\text{pairwise}({\cal C}, \widehat{\cal C}) = \frac{|p({\cal C}) \cap p(\widehat{\cal C})|}{|p({\cal C})|}\\
F_\text{pairwise}({\cal C}, \widehat{\cal C}) &= \frac{2 P_\text{pairwise}({\cal C}, \widehat{\cal C}) R_\text{pairwise}({\cal C}, \widehat{\cal C})}{P_\text{pairwise}({\cal C}, \widehat{\cal C}) + R_\text{pairwise}({\cal C}, \widehat{\cal C})}
\end{align}
\end{small}
and where $c(s)$ (resp. $\widehat{c}(s)$) is the cluster $c \in {\cal C}$ such that
$s \in c$ (resp. the cluster $\widehat{c} \in \widehat{\cal C}$ such that $s
\in \widehat{c}$), and where $p({\cal C}) = \cup_{c \in {\cal C}} \{ (s_1, s_2)
| s_1, s_2 \in c, s_1 \neq s_2 \}$ is the set of all pairs of signatures from
the same clusters in ${\cal C}$.
The F-measure is the harmonic mean between these two quantities.
In the analysis below, we rely primarily on the B3 F-measure for discussing results, as the pairwise variant
tends to favor large clusters (because the number of pairs is quadratic with the cluster size),
hence unfairly giving preference to authors with many publications.
By contrast, the B3 F-measure weights clusters linearly with respect to their size.
General conclusions drawn below remain however consistent for pairwise F.

@@

\textit{Baseline.} The simplest baseline against which we compare our results
consists in grouping all signatures sharing the same (normalized) surname(s) and
the same (normalized) first given name initial. It provides a simple and fast
solution yielding decent results, as reported at the top of
Table~\ref{table:results}

@@

\begin{itemize}
  \item Blocking: same surname and the same first given name initial strategy (SFI) and with name normalization;
  \item Linkage function: all 22 features defined in Table~\ref{table:features},
  gradient boosted regression trees as supervised learning algorithm
  and a training set of pairs built from $({\cal S}_\text{train}^\prime, {\cal C}_\text{train}^\prime)$, by balancing easy and difficult cases.
  \item Clustering: agglomerative clustering using average linkage and
  block cuts found to maximize $F_\text{B3}({\cal C}_\text{train}^\prime, \widehat{\cal C}_\text{train}^\prime, {\cal S}_\text{train}^\prime)$.
\end{itemize}

@@

Finally, as discussed previously, the step of
normalizing author names (stripping accents, removing affixes), as performed in the
state-of-the-art, is shown to be important. Results from Table~\ref{table:results} clearly suggest that not
normalizing significantly reduces performance (yielding an F-measure of $0.9830$ when normalizing,
but decreasing to $0.9791$ when raw author name strings are used instead).
